On the number of tiles visited by a\\ line segment on a rectangular grid

### 2021-08-24 paper.tex
Results on probability of maximum

\section{Square grid with real-valued lengths}

***

>>>>

\subsection{Square grid with real-valued lengths}

Consider a square grid with spacing $a \in \mathbb R^+$. For a random segment with length $\len \in \mathbb R^+$, let the function $\probmax: \mathbb R^+ \to [0,1]$ be defined such that $\probmax(\len)$ gives the probability that the segment visits the maximum possible number of tiles, $\funt(\len)$. This function is characterized by the next theorem.

For $u, v \in \mathbb N$, $\genvar \in \mathbb R^+$, let
\begin{equation}
\label{eq: g def}
g(\genvar, u, v) = \int_{\arcsin \frac u {2r}}^{\arccos \frac u {2r}} \left( r \sin \theta - \frac u 2 \right) \left( r \cos \theta - \frac v 2 \right) \diff \theta.
\end{equation}
Using the identifies $\cos^2 \alpha = 1-\sin^2 \alpha$ and $\cos(2\alpha) = 2\cos^2\alpha-1 = 1 - 2\sin^2\alpha$, the integral in \eqref{eq: g def} is computed as
\begin{equation}
\label{eq: g fin}
\begin{split}
g(\genvar, u, v) &= \frac 1 {2\pi} \biggl(
\left(\arccos\frac{u}{2\genvar}-\arcsin\frac{v}{2\genvar}\right) u v + 2\genvar^2 + \frac{u^2} 2 + \frac{v^2} 2 \\
& \quad  - u \sqrt{4\genvar^2-v^2} - v \sqrt{4\genvar^2-u^2} \biggr).
\end{split}
\end{equation}

\begin{theorem}
\label{theo: probmax, sq}
Given $a, \len \in \mathbb R^+$, consider a square grid with spacing $a$ and a uniformly random segment of length $\len$, as previously defined. The probability $\probmax(\len)$ that the segment visits the maximum number of tiles $\funt(\len)$ is
\begin{equation}
\label{eq: probmax}
\begin{split}
\probmax(\len) & = g\left(\frac{\len} a, 
\tiles-3, \tiles-3 \right) \quad\ \text{for } t \text{ odd, } \textstyle \frac \len a \in \left( \frac{\tiles-3}{\sqrt{2}}, \frac{\sqrt{(\tiles-5)^2+(\tiles-1)^2}} 2 \right]; \\
\probmax(\len) & = g\left(\frac{\len} a, \tiles-3, \tiles-3 \right) + 2 g\left(\frac{\len} a, \tiles-5, \tiles-1 \right) \\
& \text{for } t \text{ odd, } \textstyle \frac \len a \in \left (\frac{\sqrt{(\tiles-5)^2+(\tiles-1)^2}} 2,  \frac{\sqrt{(\tiles-3)^2+(\tiles-1)^2}} 2 \right]; \\
\probmax(\len) & = 2 g\left(\frac{\len} a, \tiles-4, \tiles-2 \right) \quad \text{for } t \text{ even, } \textstyle \frac \len a \in \left( \frac{\sqrt{(\tiles-4)^2+(\tiles-2)^2}} 2, \frac{\sqrt{(\tiles-6)^2+\tiles^2}} 2 \right]; \\
\probmax(\len) & = 2 g\left(\frac{\len} a, \tiles-4, \tiles-2 \right) + 2 g\left(\frac{\len} a, \tiles-6, \tiles \right) \\
& \text{for } t \text{ even, } \textstyle \frac \len a \in \left( \frac{\sqrt{(\tiles-6)^2+\tiles^2}} 2, \frac{\tiles-2}{\sqrt{2}} \right].
\end{split}
\end{equation}
\end{theorem}

\begin{proof}
The approach is similar to that used in the proof of Proposition~\ref{prop: Pr i}, but conditioning on the segment orientation $\theta$ instead of on the location of its first endpoint.

Consider a length $\len$ with corresponding maximum number of tiles $\tiles = \funt(\len)$ (given by Corollary~\ref{cor: funt, sq, form}). The first endpoint of the segment is assumed to be in the reference tile with lower-left corner $(0,0)$. The segment visits $\tiles$ tiles almost surely if and only if its second endpoint is in a tile with lower-left corner $(i-1,j-1)$, $i,j\geq 1$, $i+j-1=\tiles$, or in a tile symmetrical to this in the other three quadrants of the plane. Figure~\ref{fig: probmax_possible_tiles} shows the reference tile, with solid line, and some of the square tiles that fulfil this condition, with dashed line, for $\tiles=5$. The desired probability can be computed by considering only the tiles in the first quadrant and multiplying by $4$.

% ***Does this figure add something, or is it better to remove it?
\begin{figure}
\centering%
\includegraphics[width=.52\textwidth]{probmax_possible_tiles}%
\caption{Tiles in which the second endpoint can be for a given number of visited tiles $\tiles$ in a square grid (example for $\tiles=5$)
}%
\label{fig: probmax_possible_tiles}%
\end{figure}%

For any $i,j$ with $i+j-1=\tiles$ and for a given orientation $\theta$, the segment goes from the reference tile to that with lower-left corner $(i-1,j-1)$ if and only if shifting the segment so that its second endpoint has coordinates $(i-1,j-1)$ results in the first endpoint being in the reference tile. It cannot be the case that the segment ``overshoots'' past this tile, because that would result in a number of visited tiles exceeding the assumed maximum $\tiles$.

\begin{figure}
\centering%
\subfigure[The second endpoint is in the tile with lower-left corner $((t-1)/2,(t-1)/2)$]{%
\label{fig: probmax_funt_odd_diag}%
\includegraphics[width=.56\textwidth]{probmax_funt_odd_diag}%
}\\%\hfill%
\subfigure[The second endpoint is in the tile with lower-left corner $((t-3)/2,(t+1)/2)$]{%
\label{fig: probmax_funt_odd_offdiag}%
\includegraphics[width=.56\textwidth]{probmax_funt_odd_offdiag}%
}%
\caption{Segment orientations and positions of the first endpoint within the reference tile for a given number $\tiles$ of visited tiles. Example for odd $\tiles$
}%
\label{fig: probmax_funt_odd}%
\end{figure}%

There is a range of possible values of $\theta$ for which the above condition is satisfied. This range corresponds to the part of the arc with solid line in Figure~\ref{fig: probmax_funt_odd}. For convenience, the angle $\theta$ is defined clockwise from the downward direction, as seen in the figure. For each $\theta$ in the allowed range, the valid positions for the first endpoint of the (shifted) segment are in the shaded rectangle, whose dimensions depend on $\theta$. The fraction of the reference tile occupied by this rectangle is the probability $\probmax_{i-1,j-1}(\len,\theta)$ that the segment goes from the reference tile to that with lower-left corner $(i-1,j-1)$ conditioned on $\theta$. The desired probability, $\probmax(\len)$, is obtained by averaging $\probmax_{i-1,j-1}(\len, \theta)$ over $\theta$ and summing over all $i-1,j-1$ that can be reached from the reference tile.

The situation is different depending on whether $\tiles$ is odd or even. Consider $\tiles$ odd first. Under this assumption, the range of lengths corresponding to $\tiles$ is given by \eqref{eq: square, odd max tiles, lengths}. For any length in this interval, there are some positions within the reference file from which the tile with lower-left coordinate $((t-1)/2, (t-1)/2)$ can be reached. Equivalently, the shifted segment with one endpoint at $((t-1)/2, (t-1)/2)$ can reach the reference tile. This is illustrated in Figure~\ref{fig: probmax_funt_odd_diag}. In addition, if 
\begin{equation}
\label{eq: probmax len odd2}
\len > \frac{\sqrt{(\tiles-5)^2 + (\tiles-1)^2}}{2}
\end{equation}
the two tiles with lower-left coordinates $((t-3)/2, (t+1)/2)$ and $((t+1)/2, (t-3)/2)$ are also reachable from the reference tile; or equivalently the shifted segment can reach the reference tile, as shown in Figure~\ref{fig: probmax_funt_odd_offdiag}. This can only happen for $\tiles \geq 5$ (for $\tiles=3$ the inequality \eqref{eq: probmax len odd2} cannot be satisfied for lengths in the range given by \eqref{eq: square, odd max tiles, lengths}). In this case, by symmetry, each of those two tiles contributes the same probability. Other tiles corresponding to the same $\tiles$ need not be considered, because they cannot be reached with lengths satisfying \eqref{eq: square, odd max tiles, lengths}.

For $\tiles$ even, which corresponds to lengths in the interval \eqref{eq: square, even max tiles, lengths}, the two tiles with lower-left coordinates $(t/2-1, t/2)$ and $(t/2, t/2-1)$ can always be reached from some positions within the reference tile. In addition, if
\begin{equation}
\label{eq: probmax len even2}
\len > \frac{\sqrt{((\tiles-6)^2 + \tiles^2}}{2}
\end{equation}
the two tiles with lower-left coordinates $(\tiles/2-2, \tiles/2+1)$ and $(\tiles/2+1, \tiles/2-2)$ have to be considered too. This can only happen for $\tiles \geq 8$. No other tiles can be reached from the reference tile.

Therefore, two cases need to be distinguished for $\tiles$ odd, depending on whether \eqref{eq: probmax len odd2} is satisfied or not; and similarly two cases exist for $\tiles$ even, depending on whether \eqref{eq: probmax len even2} holds or not. These are, respectively, the four cases in \eqref{eq: probmax}.

In the first case, as has been argued, only the tile with lower-left coordinates $((t-1)/2, (t-1)/2)$ is reachable. For a given $\theta$, the conditional probability $\probmax_{(t-1)/2, (t-1)/2}(\len, \theta)$ is the area of the shaded rectangle in Figure~\ref{fig: probmax_funt_odd_diag} divided by $a^2$. The horizontal and vertical sides of this rectangle are $\len\sin\theta - a(\tiles-3)/2$ and $\len\cos\theta - a(\tiles-3)/2$ respectively, and thus
\begin{equation}
\label{eq: probmax: probmax theta odd}
\probmax_{(t-1)/2, (t-1)/2}(\len, \theta) = \frac{(\len\sin\theta - a(\tiles-3)/2) (\len\cos\theta - a(\tiles-3)/2)} {a^2}
\end{equation}
The range of allowed values for $\theta$, as can be seen from Figure~\ref{fig: probmax_funt_odd_diag}, is $(\theta_0, \theta_1)$ with
\begin{align}
\label{eq: odd theta 0}
\theta_0 & = \arcsin\frac{a(\tiles-3)}{2\len}, \\
\label{eq: odd theta 1}
\theta_1 & = \arccos\frac{a(\tiles-3)}{2\len}.
\end{align}
Computing the average of \eqref{eq: probmax: probmax theta odd} over $\theta$, multiplying by $4$ to include all quadrants and comparing with \eqref{eq: g def} gives $\probmax(\len)$ as
\begin{equation}
\label{eq: probmax: probmax odd}
\begin{split}
\probmax(\len) & = 4 \int_{\theta_0}^{\theta_1} \frac{\probmax_{(t-1)/2, (t-1)/2}(\len, \theta)}{2\pi} \diff \theta \\
& = \frac 2 \pi \int_{\arcsin\frac{a(\tiles-3)}{2\len}}^{\arccos\frac{a(\tiles-3)}{2\len}} \left( \frac{\len\sin\theta}a - \frac{\tiles-3} 2 \right)  \left( \frac{\len\cos\theta}a - \frac{\tiles-3} 2 \right) \diff \theta \\
& = g\left( \frac \len a, t-3, t-3 \right ),
\end{split}
\end{equation}
in accordance with \eqref{eq: probmax}.

For the second case in \eqref{eq: probmax}, $\probmax(\len)$ is obtained as in the preceding paragraph plus an additional term corresponding to the contribution of the tile with lower-left corner $((t-3)/2, (t+1)/2)$, multiplied by $2$ to account for its symmetrical tile $((t+1)/2, (t-3)/2)$ and by $4$ to include all quadrants. The sides of the shaded rectangle in Figure~\ref{fig: probmax_funt_odd_offdiag} are $\len\sin\theta - a(\tiles-5)/2$ and $\len\cos\theta - a(\tiles-1)/2$, and therefore
\begin{equation}
\label{eq: probmax: probmax theta odd2}
\probmax_{(t-3)/2, (t+1)/2}(\len, \theta) = \frac{(\len\sin\theta - a(\tiles-5)/2) (\len\cos\theta - a(\tiles-1)/2)} {a^2}.
\end{equation}
The integration interval $(\theta_0, \theta_1)$ is in this case
\begin{align}
\theta_0 & = \arcsin\frac{a(\tiles-5)}{2\len}, \\
\theta_1 & = \arccos\frac{a(\tiles-1)}{2\len}.
\end{align}
The additional contribution to $\probmax(\len)$, given by
\begin{multline}
8 \int_{\theta_0}^{\theta_1} \frac{\probmax_{(t-3)/2, (t+1)/2}(\len, \theta)}{2\pi} \diff \theta \\
= \frac 4 \pi \int_{\arcsin\frac{a(\tiles-5)}{2\len}}^{\arccos\frac{a(\tiles-1)}{2\len}} \left( \frac{\len\sin\theta}a - \frac{\tiles-5} 2 \right)  \left( \frac{\len\cos\theta}a - \frac{\tiles-1} 2 \right) \diff \theta,
\end{multline}
is recognized to be $2 g(\len/a, \tiles-5, \tiles-1)$. This proves the second equality in \eqref{eq: probmax}.

The expressions for the third and fourth cases in \eqref{eq: probmax}, corresponding to $\tiles$ even, are established analogously.
\end{proof}

Figure~\ref{fig: probmax_len} shows the probability $\probmax(\len)$ of visiting the maximum number of tiles on a square grid with spacing $a$, as given by \eqref{eq: probmax}. As $\len$ grows, $\probmax(\len)$ has a jump when the maximum number of tiles that can be visited increases by $1$. This happens when $\len$ equals the right endpoint of the interval \eqref{eq: square, odd max tiles, lengths} for $\tiles$ odd or of \eqref{eq: square, even max tiles, lengths} for $\tiles$ even. Let these length values be denoted as $\len_\tiles$:
\begin{equation}
\len_\tiles = \begin{cases}
\displaystyle
\frac{a \sqrt{(\tiles-3)^2 + (\tiles-1)^2}} {2} & \text{ for $\tiles$ odd, $\tiles \geq 3$} \\
\displaystyle
\frac{a (\tiles-2)}{\sqrt{2}} & \text{ for $\tiles$ odd, $\tiles \geq 4$}.
\end{cases}
\end{equation}
For example, the first continuous section seen in the figure corresponds to a maximum number of tiles $\tiles=3$, for lengths in the interval $(0, \len_3]$, with $\len_3= a$. The second corresponds to $\tiles=4$, for lengths in $(\len_3, \len_4]$, with $\len_4= a\sqrt{2}$. Within each continuous section the probability monotonically increases from $0$ to a maximum value. The heights of the maxima follow a different trend*** for odd and even $\tiles$, as seen in the figure; but in each case they are asymptotically proportional to $1/\len_\tiles$, as the next result establishes.

\begin{figure}%
\centering%
\includegraphics[width=.75\textwidth]{probmax_len}%
\caption{Probability $\probmax(\len)$ that a random segment of length $\len$ visits the maximum number of tiles%
}%
\label{fig: probmax_len}%
\end{figure}%

\begin{proposition}
The function $\probmax$ has the following asymptotic behaviour:
\begin{align}
\label{eq: probmax liminf}
\liminf_{\len \rightarrow \infty}\, \probmax(\len) & = 0 \\ 
\label{eq: probmax lim odd}
\lim_{\substack{\tiles \rightarrow \infty \\
\tiles \ \mathrm{odd}}}\, \tiles \probmax(\len_\tiles) & = \frac 2 \pi \\
\label{eq: probmax lim even}
\lim_{\substack{\tiles \rightarrow \infty \\ \tiles \ \mathrm{even}}}\, \tiles \probmax(\len_\tiles) & = \frac 8 {3\pi} \\
\label{eq: probmax limsup}
\limsup_{\len \rightarrow \infty}\, \len \probmax(\len) & = \frac {4\sqrt{2} a}{3\pi}.
\end{align}
\end{proposition}

\begin{proof}
The probability $\probmax(\len)$ becomes $0$ whenever the length causes the maximum number of visited tiles to increase by $1$. This establishes \eqref{eq: probmax liminf}.

For \eqref{eq: probmax lim odd} it suffices to note that
\begin{align}
\label{eq: tedious limit, odd 1}
\lim_{\tiles \rightarrow \infty}\, \tiles g\left(\frac{\sqrt{(\tiles-3)^2+(\tiles-1)^2}} 2, \tiles-3, \tiles-3 \right) &= \frac 2 {3\pi} \\
\label{eq: tedious limit, odd 2}
\lim_{\tiles \rightarrow \infty}\, \tiles g\left(\frac{\sqrt{(\tiles-3)^2+(\tiles-1)^2}} 2, \tiles-5, \tiles-1 \right)
& = \frac 2 {3\pi},
\end{align}
as can be checked using L'H\^opital's rule \cite[section~5.3]{Abbott15}.
%***I think. It's very tedious by hand. I used Maple to compute the limit, and then checked it numerically. Maybe one or two intermediate steps from applying L'H\^opital's rule should be included. Perhaps those can be obtained with Maple too, and this part can be rewritten.
Combining these expressions with the second equality in \eqref{eq: probmax} yields \eqref{eq: probmax lim odd}.

Similarly,
\begin{align}
\label{eq: tedious limit, even 1}
\lim_{\tiles \rightarrow \infty}\, \tiles g\left(\frac{\tiles-2}{\sqrt 2}, \tiles-4, \tiles-2 \right) &= \frac 2 {3\pi} \\
\label{eq: tedious limit, even 2}
\lim_{\tiles \rightarrow \infty}\, \tiles g\left(\frac{\tiles-2}{\sqrt 2}, \tiles-6, \tiles \right)
& = \frac 2 {3\pi},
\end{align}
and combining with the fourth equality in \eqref{eq: probmax} gives \eqref{eq: probmax lim even}.

For large $\tiles$, $\len_\tiles$ asymptotically approaches $a\tiles/\sqrt{2}$. Using this into \eqref{eq: probmax lim even}, which is greater than \eqref{eq: probmax lim odd}, establishes \eqref{eq: probmax limsup}.
\end{proof}
